### 2012-09-23 paper_sean/conclusion.tex

% !TEX root =main.tex
HyperNEAT has proven to be an effective encoding in gait evolution with an ability to find gait regularities without human intervention \cite{clune1,clune2,yos:clune}. 
Glette et al. \cite{glette} has shown that evolving gaits in simulation using a simple algorithm with a large number of evaluations could outperform gait evolution using a more sophisticated algorithm in hardware thanks to the simulator affording a larger number of evaluations to be performed for evolution. 
In this study, we took the two -- HyperNEAT and a physical simulator -- and showed that HyperNEAT, when it could afford many evaluations, could evolve impressive gaits. 
In simulation, HyperNEAT gaits outperformed Genetic Algorithm gaits \cite{glette} by 54.3\% and by 5.4\% in the real life trials. 
This study also showed that HyperNEAT will exploit the characteristics of the problem, such as ferquency and regularity, in order to maximize its fitness.
%Without frequency limiters, it would make the legs move really fast in order to increase fitness. 
%When that was no longer allowed, it pushed the servo positions to their boundaries by increasing the stride lengths as wide as possible. 
However, because of the physical limitations of the robot, these exploitations did not transfer well to reality. %could not be transferred well to reality.
Moreover, the inevitable faults with the simulator (it cannot model the robot or the real world perfectly) added to the difficulty in the transfer from simulation to reality. 
Thus, many of the gaits performed much worse than it did in simulation and had difficulty in achieving consistent results.

Next steps for this research heads towards improving the transfer process of the gaits from simulation to reality. 
The performance gap between simulation and reality is called the \emph{reality gap} \cite{jakobi}. 
Many different approaches to reduce the reality gap has been proposed \cite{koos1,bongard,zagal}. 
To improve upon the results from this study, one or more of these approaches could be used to aid in the transfer to reality. 
The one easiest and most promising to implement is the Transferability Approach \cite{koos2} which has shown to be successful in quadrapedal gait evolution. 
HyperNEAT, used with a simulator and a transfer approach, shows much hope in producing yet a better, more coordinated, and higher performing gait on the QuadraTot platform.

>>>>

% !TEX root =main.tex
HyperNEAT has proven to be an effective encoding in gait evolution with an ability to find gait regularities without human intervention \cite{clune2009evolving,clune2011performance,yos:clune}. 
Glette et al. \cite{glette} has shown that evolving gaits in simulation using a simple algorithm with a large number of evaluations could outperform gait evolution using a more sophisticated algorithm in hardware thanks to the simulator affording a larger number of evaluations to be performed for evolution. 
In this study, we took the two -- HyperNEAT and a physical simulator -- and showed that HyperNEAT, when it could afford many evaluations, could evolve impressive gaits. 
In simulation, HyperNEAT gaits outperformed Genetic Algorithm gaits \cite{glette} by 54.3\% and by 5.4\% in the real life trials. 
This study also showed that HyperNEAT will exploit the characteristics of the problem, such as ferquency and regularity, in order to maximize its fitness.
%Without frequency limiters, it would make the legs move really fast in order to increase fitness. 
%When that was no longer allowed, it pushed the servo positions to their boundaries by increasing the stride lengths as wide as possible. 
However, because of the physical limitations of the robot, these exploitations did not transfer well to reality. %could not be transferred well to reality.
Moreover, the inevitable faults with the simulator (it cannot model the robot or the real world perfectly) added to the difficulty in the transfer from simulation to reality. 
Thus, many of the gaits performed much worse than it did in simulation and had difficulty in achieving consistent results.

Next steps for this research heads towards improving the transfer process of the gaits from simulation to reality. 
The performance gap between simulation and reality is called the \emph{reality gap} \cite{jakobi}. 
Many different approaches to reduce the reality gap has been proposed \cite{koos2010crossing,bongard,zagal}. 
To improve upon the results from this study, one or more of these approaches could be used to aid in the transfer to reality. 
The one easiest and most promising to implement is the Transferability Approach \cite{koos2011transferability} which has shown to be successful in quadrapedal gait evolution. 
HyperNEAT, used with a simulator and a transfer approach, shows much hope in producing yet a better, more coordinated, and higher performing gait on the QuadraTot platform.

### 2012-09-23 paper_sean/hyperneat.tex

% !TEX root =main.tex
HyperNEAT is a neuroevolution method that evolves artificial neuron networks (ANNs) using evolutionary algorithms.
HyperNEAT has shown to evolve large neural networks that represent many features present in brains such as repetitions and regularities \cite{stanley3}. 
This encoding has previously been applied for gait-learning \cite{yos:clune} and for evolving 2-D and 3-D objects \cite{clune:lipson} and has been successful in displaying complex, natural characteristics such as symmetries and repetition of themes often found in nature.

HyperNEAT indirectly encodes ANNs using compositional pattern producing networks (CPPNs) which represent genomes encoding ANN phenotypes \cite{stanley1}, \cite{stanley2}. 
The motivation for CPPNs is that complex patterns can be produced by determining the attributes of their phenotypic components as a function of their geometric location. 
This is based on the belief that in nature, cells in a body differentiates and specializes in different things with respect to their location. 
In real life, a cell in natural organisms cannot determine its location in space without the help of chemical gradients but \emph{in silico}, cells can be given their geometric coordinates, which CPPNs exploit. 
%A cell in natural organisms cannot determine its location in space by itself and requires chemical gradients are used to signal its location \cite{carroll}. 
%In contrast, \emph{in silico}, cells can be given their geometric coordinates and CPPNs use this to their advantage.

Each CPPN is a directed graph in which every node is a mathematical function, such as sine or Gaussian. 
Moreover, a CPPN genome allows functions to be made of other functions, allowing coordinate frames to be combined and hierarchies to develop, allowing for several different themes to appear such as a repeating theme (sine) with symmetry (Gaussian). 
%For example, a sine function could develop a repeating theme which is passed onto a Gaussian function to create a repeating series of symmetrical motifs. 
These functions allow evolution to exploit various properties that allow HyperNEAT to find regularities without further human intervention \cite{stanley1,clune1}. %such as symmetry (e.g. Gaussian) and repetition (e.g. sine function) \cite{stanley1}. 
%Because evolution is able to exploit such properties, things like gaits are able to feature repetition and symmetry on its own without human intervention \cite{clune1}.

@@

HyperNEAT is first neuroevolutionary algorithm that has shown capabilities in exploiting the geometry of the problem \cite{stanley1}, \cite{clune3}.
This is possible because the connection weights are a function of geometric positions of these nodes. 
If positions inputed to CPPNs represent aspects of the problem relevant to the solution, HyperNEAT could use that information to its advantage.

HyperNEAT evolves the CPPNs according to the principles of the Neuroevolution of Augmenting Topologies (NEAT) algorithm \cite{stanley4}. 
The NEAT algorithm has three major components \cite{stanley4}. 
First, it starts with small genomes that encode simple networks and complexifies them via mutations and add nodes and links to the network which allows the alogithm to evolve the network topology and weight. 
Second, NEAT has a fitness-sharing system that preserves diversity in the system and allows for new innovations to be tuned by evolution before competing them against more adapted rivals. 
Third, the historical information stored in genes helps to perform crossover in a way that is effective, yet avoids the need for expensive topological analysis. A full explanation of NEAT can be found in Stanley and Miikkulainen \cite{stanley4}.

@@

% \begin{figure}
% \begin{center}
% \vspace{1cm}
% \epsfig{file=CPPN.eps, width=8cm}
% \caption [ ]{Compositional Pattern Producing Networks (CPPNs). CPPN  genomes use mathematical functions to generate regularities such as symmetry and repetitions. CPPN genomes allow functions to be made of other functions, allowing multiple regularity motifs to be present. Figure taken from Stanley \cite{stanley2}.}
% \end{center}
% \end{figure}
\figp{hyperneat_bug_example}{.6}{Compositional Pattern Producing Networks (CPPNs). CPPN  genomes use mathematical functions to generate regularities such as symmetry and repetitions. CPPN genomes allow functions to be made of other functions, allowing multiple regularity motifs to be present. Figure taken from Stanley \cite{stanley2}.}

% \begin{figure}
% \begin{center}
% \vspace{1cm}
% \epsfig{file=hyperneat.eps, width=8cm}
% \caption [ ]{Producing ANNs from CPPNs. In HyperNEAT, weights are specified as a function of the Cartesian coordinates for source and input nodes and a constant bias for each connection between the source and input. All pairwise combinations of source and target node coordinates are iteratively passed into CPPNs to determine the weight of each ANN link. Figure taken from Clune et al. \cite{clune2}}.
% \end{center}
% \end{figure}
\figp{hyperneatExplanation}{.6}{Producing ANNs from CPPNs. In HyperNEAT, weights are specified as a function of the Cartesian coordinates for source and input nodes and a constant bias for each connection between the source and input. All pairwise combinations of source and target node coordinates are iteratively passed into CPPNs to determine the weight of each ANN link. Figure taken from Clune et al. \cite{clune2}}

>>>>

% !TEX root =main.tex
HyperNEAT is a neuroevolution method that evolves artificial neuron networks (ANNs) using evolutionary algorithms.
HyperNEAT has shown to evolve large neural networks that represent many features present in brains such as repetitions and regularities \cite{stanley2002evolving}. 
This encoding has previously been applied for gait-learning \cite{yos:clune} and for evolving 2-D and 3-D objects \cite{clune:lipson20113d} and has been successful in displaying complex, natural characteristics such as symmetries and repetition of themes often found in nature.

HyperNEAT indirectly encodes ANNs using compositional pattern producing networks (CPPNs) which represent genomes encoding ANN phenotypes \cite{stanley2009hypercube}, \cite{stanley2007CPPN}. 
The motivation for CPPNs is that complex patterns can be produced by determining the attributes of their phenotypic components as a function of their geometric location. 
This is based on the belief that in nature, cells in a body differentiates and specializes in different things with respect to their location. 
In real life, a cell in natural organisms cannot determine its location in space without the help of chemical gradients but \emph{in silico}, cells can be given their geometric coordinates, which CPPNs exploit. 
%A cell in natural organisms cannot determine its location in space by itself and requires chemical gradients are used to signal its location \cite{carroll}. 
%In contrast, \emph{in silico}, cells can be given their geometric coordinates and CPPNs use this to their advantage.

Each CPPN is a directed graph in which every node is a mathematical function, such as sine or Gaussian. 
Moreover, a CPPN genome allows functions to be made of other functions, allowing coordinate frames to be combined and hierarchies to develop, allowing for several different themes to appear such as a repeating theme (sine) with symmetry (Gaussian). 
%For example, a sine function could develop a repeating theme which is passed onto a Gaussian function to create a repeating series of symmetrical motifs. 
These functions allow evolution to exploit various properties that allow HyperNEAT to find regularities without further human intervention \cite{stanley2009hypercube,clune2009evolving}. %such as symmetry (e.g. Gaussian) and repetition (e.g. sine function) \cite{stanley1}. 
%Because evolution is able to exploit such properties, things like gaits are able to feature repetition and symmetry on its own without human intervention \cite{clune1}.

@@

HyperNEAT is first neuroevolutionary algorithm that has shown capabilities in exploiting the geometry of the problem \cite{stanley2009hypercube}, \cite{clune2009sensitivity}.
This is possible because the connection weights are a function of geometric positions of these nodes. 
If positions inputed to CPPNs represent aspects of the problem relevant to the solution, HyperNEAT could use that information to its advantage.

HyperNEAT evolves the CPPNs according to the principles of the Neuroevolution of Augmenting Topologies (NEAT) algorithm \cite{stanley2006exploiting}. 
The NEAT algorithm has three major components \cite{stanley2006exploiting}. 
First, it starts with small genomes that encode simple networks and complexifies them via mutations and add nodes and links to the network which allows the alogithm to evolve the network topology and weight. 
Second, NEAT has a fitness-sharing system that preserves diversity in the system and allows for new innovations to be tuned by evolution before competing them against more adapted rivals. 
Third, the historical information stored in genes helps to perform crossover in a way that is effective, yet avoids the need for expensive topological analysis. A full explanation of NEAT can be found in Stanley and Miikkulainen \cite{stanley2006exploiting}.

@@

% \begin{figure}
% \begin{center}
% \vspace{1cm}
% \epsfig{file=CPPN.eps, width=8cm}
% \caption [ ]{Compositional Pattern Producing Networks (CPPNs). CPPN  genomes use mathematical functions to generate regularities such as symmetry and repetitions. CPPN genomes allow functions to be made of other functions, allowing multiple regularity motifs to be present. Figure taken from Stanley \cite{stanley2}.}
% \end{center}
% \end{figure}
\figp{hyperneat_bug_example}{.6}{Compositional Pattern Producing Networks (CPPNs). CPPN  genomes use mathematical functions to generate regularities such as symmetry and repetitions. CPPN genomes allow functions to be made of other functions, allowing multiple regularity motifs to be present. Figure taken from Stanley \cite{stanley2007CPPN}.}

% \begin{figure}
% \begin{center}
% \vspace{1cm}
% \epsfig{file=hyperneat.eps, width=8cm}
% \caption [ ]{Producing ANNs from CPPNs. In HyperNEAT, weights are specified as a function of the Cartesian coordinates for source and input nodes and a constant bias for each connection between the source and input. All pairwise combinations of source and target node coordinates are iteratively passed into CPPNs to determine the weight of each ANN link. Figure taken from Clune et al. \cite{clune2}}.
% \end{center}
% \end{figure}
\figp{hyperneatExplanation}{.6}{Producing ANNs from CPPNs. In HyperNEAT, weights are specified as a function of the Cartesian coordinates for source and input nodes and a constant bias for each connection between the source and input. All pairwise combinations of source and target node coordinates are iteratively passed into CPPNs to determine the weight of each ANN link. Figure taken from Clune et al. \cite{clune2011performance}}.

### 2012-09-23 paper_sean/introduction.tex
small changes
ridder storm bib add

% !TEX root =main.tex
Robots that can move around on legs can operate in a much wider range of environments than their wheeled counterparts. A downside is that designing gaits for legged robots is a difficult and time-consuming process~\cite{ridder storm, wetter green}. 
But designing a gait for a robot is often difficult and time consuming because the engineer must take into consideration many factors which must all fall into place for a successful gait pattern. 
Because of setbacks present in hand-designing gaits, evolving gaits with machine learning algorithms is an attractive and flexible approach that can automatically produce effective gaits for many different robot platforms with minimal human intervention.

Much research has been done in this subject, and previous work has shown that learned gaits can outperform designed gaits \cite{valsalam:mii,kohl:stone,hornby1,hornby2}. %JMC/SL- I have a similar claim in my CEC 2009 paper. Grab those cites for here (especially the Hornby ones). / have added hornby citations
It has also been shown that gaits with regularity (coordination in leg movement, such as left-right symmetry or front-back symmetry) perform better than gaits without regularities \cite{valsalam:mii,clune2,clune1,clune3}. %JMC/SL-note how you can put multiple cites in one bracket...but no spaces after the comma! / gotcha!
However, in most previous works, the experimenter has had to manually decide the regularities of the gaits \cite{valsalam:mii,tellez,beer,raibert} . %JMC/SL - we need cites for this claim. you can get them from Clune1 / done
Such manual intervention is time consuming, requires expert knowledge, and adds constraints that may hurt performance. 
Previous work has shown that the Hypercube-based NeuroEvolution of Augmenting Topologies (HyperNEAT) generative encoding \cite{stanley1} can automatically generate a variety of regular gaits and that it outperforms direct encoding controls on this task \cite{clune1}, \cite{clune2}. 
However, that work only verified these claims in simulation. 
A follow-up paper by Yosinski et al. evolved gaits with HyperNEAT directly in hardware on the QuadraTot robot platform and found that HyperNEAT's gaits outperformed manually designed parameterized learning algorithms, but still did not produce impressive, natural gaits \cite{yos:clune}. 
A further follow-up study built a simulator for the QuadraTot robot platform to test whether the inclusion of a simulator would improve results and found that it did:  an evolutionary algorithm with direct encoding in simulation produced gaits that outperformed those gaits evolved by HyperNEAT in hardware \cite{glette}. %JMC/SL-Should there be a space between the end of the sentence and the cite? Whatever the answer, we should be consistent. If so, I have added here with the ~. The ~ tells Latex not to split that space from the character before and after it. / There should have been and will make changes accordingly
The simulator reduced the time required to evaluate gaits and afforded much larger population sizes and more generations than was possible when evolving in hardware, ultimately performing 333 times as many evaluations as HyperNEAT had in hardware per run (60000 vs. 180) \cite{yos:clune}, \cite{glette}. 
The evolutionary algorithm used in Glette et al. \cite{glette} was manually constrained to evolve specific, hand-chosen regularities. 
%Another study evolved gaits in hardware on the QuadraTot platform using RL PoWER, a spline-based algorithm, and was successful in evolving faster and more repeatable gaits than HyperNEAT evolved in hardware \cite{haocheng}. %JMC- I don't think we need to cite this paper...it is a non-evolutionary method, and it doesn't outperform Glette...and the sentence just sticks out as a detail that isn't necessary to our story.

In this paper we test whether the benefits of HyperNEAT -- higher performance and the automatic discovery of complex, effective regularities \cite{clune2,yos:clune} -- will outperform a simple genetic algorithm manually designed to produce regularities once HyperNEAT is combined with a simulator. %JMC/SL-I don't think they used a direct encoding. I prefer describing it as simple and manually designed to produce specific regularities. / changed to reflect that
We test this by evolving gaits using HyperNEAT with a physical simulator and then comparing the performance of these gaits in the real-world to those that were evaluated on the QuadraTot robotic platform in previous publications.  %JMC - Try to write without ever saying ``use'' or ``using.'' If you do, you'll find how much it is overused in writing, as evidenced by this last sentence. Often the word is a stand in for a different, clearer verb...and its use is ambiguous. Please search for this word and try to swap it out throughout the rest of the paper. Often times it requires restructuring a sentence, but the sentence usually ends up better. Similarly, make sure to always follow the word ``this'' with referent. This is extremely ambiguous. You know what you refer to when you write the sentence, but readers (often including your future self!) do not know precisely what you mean, and in science precision is critical. 
%Our experiments confirm the hypothesis: gaits evolved with HyperNEAT and then transferred to reality were the highest performing observed to date on the QuadraTot platform. 
%That said, gaits still transferred poorly, revealing that substantial further performance gains are likely via future work that combines a simulator, HyperNEAT, and techniques that reduce the simulator-to-reality gap. 
%More broadly, this work supports previous work that evolving with a simulator and then transferring to reality is a more powerful means of achieving results than evolving directly in hardware. 
%Our work also confirms that generative encodings such as HyperNEAT are a powerful way to evolve gaits and, more generally, automatically exploit regularities in challenging engineering domains.

>>>>

% !TEX root =main.tex
Robots that can move around on legs can operate in a much wider range of environments than their wheeled counterparts. A downside is that designing gaits for legged robots is a difficult and time-consuming process for human engineers~\cite{strom1999legged,wettergreen1992gait}. 
But designing a gait for a robot is often difficult and time consuming because the engineer must take into consideration many factors which must all fall into place for a successful gait pattern. 
Because of setbacks present in hand-designing gaits, evolving gaits with machine learning algorithms is an attractive and flexible approach that can automatically produce effective gaits for many different robot platforms with minimal human intervention.

Much research has been done in this subject, and previous work has shown that learned gaits can outperform designed gaits \cite{valsalam:mii,kohl:stone,hornby2005autonomous,hornby2003generative}. 
It has also been shown that gaits with regularity (coordination in leg movement, such as left-right symmetry or front-back symmetry) perform better than gaits without regularities \cite{valsalam:mii,clune2011performance,clune2009evolving,clune2009sensitivity}. %JMC/SL-note how you can put multiple cites in one bracket...but no spaces after the comma! / gotcha!
However, in most previous works, the experimenter has had to manually decide the regularities of the gaits \cite{valsalam:mii,tellez,beer,raibert} . %JMC/SL - we need cites for this claim. you can get them from Clune1 / done
Such manual intervention is time consuming, requires expert knowledge, and adds constraints that may hurt performance. 
Previous work has shown that the Hypercube-based NeuroEvolution of Augmenting Topologies (HyperNEAT) generative encoding \cite{stanley2009hypercube} can automatically generate a variety of regular gaits and that it outperforms direct encoding controls on this task \cite{clune2009evolving}, \cite{clune2011performance}. 
However, that work only verified these claims in simulation. 
A follow-up paper by Yosinski et al. evolved gaits with HyperNEAT directly in hardware on the QuadraTot robot platform and found that HyperNEAT's gaits outperformed manually designed parameterized learning algorithms, but still did not produce impressive, natural gaits \cite{yos:clune}. 
A further follow-up study built a simulator for the QuadraTot robot platform to test whether the inclusion of a simulator would improve results and found that it did:  an evolutionary algorithm with direct encoding in simulation produced gaits that outperformed those gaits evolved by HyperNEAT in hardware \cite{glette}. %JMC/SL-Should there be a space between the end of the sentence and the cite? Whatever the answer, we should be consistent. If so, I have added here with the ~. The ~ tells Latex not to split that space from the character before and after it. / There should have been and will make changes accordingly
The simulator reduced the time required to evaluate gaits and afforded much larger population sizes and more generations than was possible when evolving in hardware, ultimately performing 333 times as many evaluations as HyperNEAT had in hardware per run (60000 vs. 180) \cite{yos:clune}, \cite{glette}. 
The evolutionary algorithm used in Glette et al. \cite{glette} was manually constrained to evolve specific, hand-chosen regularities. 
%Another study evolved gaits in hardware on the QuadraTot platform using RL PoWER, a spline-based algorithm, and was successful in evolving faster and more repeatable gaits than HyperNEAT evolved in hardware \cite{haocheng}. %JMC- I don't think we need to cite this paper...it is a non-evolutionary method, and it doesn't outperform Glette...and the sentence just sticks out as a detail that isn't necessary to our story.

In this paper we test whether the benefits of HyperNEAT -- higher performance and the automatic discovery of complex, effective regularities \cite{clune2011performance,yos:clune} -- will outperform a simple genetic algorithm manually designed to produce regularities once HyperNEAT is combined with a simulator. %JMC/SL-I don't think they used a direct encoding. I prefer describing it as simple and manually designed to produce specific regularities. / changed to reflect that
We test this by evolving gaits using HyperNEAT with a physical simulator and then comparing the performance of these gaits in the real-world to those that were evaluated on the QuadraTot robotic platform in previous publications.  %JMC - Try to write without ever saying ``use'' or ``using.'' If you do, you'll find how much it is overused in writing, as evidenced by this last sentence. Often the word is a stand in for a different, clearer verb...and its use is ambiguous. Please search for this word and try to swap it out throughout the rest of the paper. Often times it requires restructuring a sentence, but the sentence usually ends up better. Similarly, make sure to always follow the word ``this'' with referent. This is extremely ambiguous. You know what you refer to when you write the sentence, but readers (often including your future self!) do not know precisely what you mean, and in science precision is critical. 
%Our experiments confirm the hypothesis: gaits evolved with HyperNEAT and then transferred to reality were the highest performing observed to date on the QuadraTot platform. 
%That said, gaits still transferred poorly, revealing that substantial further performance gains are likely via future work that combines a simulator, HyperNEAT, and techniques that reduce the simulator-to-reality gap. 
%More broadly, this work supports previous work that evolving with a simulator and then transferring to reality is a more powerful means of achieving results than evolving directly in hardware. 
%Our work also confirms that generative encodings such as HyperNEAT are a powerful way to evolve gaits and, more generally, automatically exploit regularities in challenging engineering domains.
